Estimation of flow trajectories in a multi-lines network

### 2023-02-10 latex_files/applied_network_science_article/article_v0.tex
Finished (for now) with chapter2: Notations and formalism

Let  $i=1,\ldots, l$ enumerate the bus stops in increasing order,  i.e. $F(i)=i+1$. The initial prior is simply $g_{st}=\frac{I(s<t)}{(l-1)(l-2)}$ (where $I(.)$ denotes the 0/1  indicator function), reflecting only the unidirectional nature of trips. The margins of the empirical distribution $f_{st}$ as well as the total flow are know, namely 
\begin{displaymath}
\alpha_s=\frac{a_s}{a_\bullet}\qquad\qquad\qquad \beta_t=\frac{b_t}{b_\bullet}\qquad\qquad\qquad n_{\bullet\bullet}=a_{\bullet}=b_\bullet
\end{displaymath}
and (\ref{Sol}) yields, after reparameterization, maximum entropy flows of the form
\begin{equation}
\label{nosignle}
n_{st}=I(s<t)\, c_s\, d_t \qquad\qquad\mbox{where}\quad \sum_{s}c_s\, \sum_{t>s}d_t=n_{\bullet\bullet}
\end{equation}
where the constraints (\ref{Sol_LagMult}) equivalently express as 
\begin{equation}
\label{dis embarking constraints}
c_s=\frac{a_s}{\sum_{t>s}d_t}
\qquad\qquad\qquad
d_t=\frac{b_t}{\sum_{s<t}c_s}
\end{equation}
to be solved by iterative fitting.

The form (\ref{nosignle}) for the flows is reminiscent of the {\em gravity flows} \cite{wilson1967statistical}  \cite{erlander1990gravity}, where 
$c_s$ is the {\em push factor}, $d_t$ is the {\em pull factor}, and $I(s<t)$ the {\em distance deterrence function}. Yet, instead of  
being symmetric in $s,t$ and decreasing with the distance $|s-t|$, the distance deterrence function is here asymmetric due to line orientation, but otherwise constant. This constancy entails the following Markovian behavior of flows: let $m_{st}$ be the number of travelers embarking on $s$ and still present at stop $t>s$, and $\rho_s$ the probability to disembark at $s$. Then, by definition and (\ref{nosignle}), ****
\begin{displaymath}
n_{st}=m_{st}\rho_t \qquad\qquad n_{s,t+1}=m_{st}(1-\rho_t)\rho_{t+1}\qquad\qquad \frac{n_{s,t+1}}{n_{st}}=\frac{(1-\rho_t)\rho_{t+1}}{\rho_t}=\frac{d_{t+1}}{d_t}
\end{displaymath}
Ainsi 
\begin{displaymath}
\frac{d_{t+2}}{d_{t+1}}=\frac{(1-\rho_{t+1})\rho_{t+2}}{\rho_{t+1}}
\qquad\qquad \frac{d_{t+2}}{d_{t}}=\frac{d_{t+2}}{d_{t+1}}\frac{d_{t+1}}{d_{t}}=\frac{(1-\rho_{t+1})\rho_{t+2}}{\rho_{t+1}}\frac{(1-\rho_t)\rho_{t+1}}{\rho_t}=
\rho_{t+2}(1-\rho_{t+1})(1-\rho_{t})\frac{1}{\rho_{t}}
\end{displaymath}
\begin{displaymath}
\frac{d_{t+q}}{d_{t}}=
\rho_{t+q}(1-\rho_{t+q-1})(1-\rho_{t+q-2})\ldots(1-\rho_{t+1}) (1-\rho_{t})\frac{1}{\rho_{t}}
\end{displaymath}
\begin{displaymath}
d_l+d_{l-1}=d_2[\rho_l(1-\rho_{l-1})+\rho_{l-1}](1-\rho_{l-2})(1-\rho_{l-3})\ldots(1-\rho_{3}) (1-\rho_{2})\frac{1}{\rho_{2}}
\end{displaymath}
\begin{displaymath}
D_i=\sum_{t>i}d_t
\end{displaymath}
Soit $z_{st}$ la proportion de voyageurs présents dans le bus en $s$ et toujours présents en $t>s$. Soit $\tau_s$ la probabilité de monter en $s$. Alors
$f_{st}=\tau_sz_{st}\rho_t$

@@

@@

*****

where $D_s=\sum_{t>s}d_t$ and $C_t=\sum_{s<t}c_s$. This maximum entropy solution constitutes one possible consistent transportation flow among many others, such as the ``first in, first out" (FIFO) flow. Interestingly enough, (\ref{nosignle}) is reminiscent of the so-called gravity flows of quantitative Geography (**ref**) $n_{st}=h_{st}\, c_s\, d_t $, but with a purely asymmetric ``distance deterrence function" $h_{st}=I(s<t)$.  Also, 
(\ref{nosignle}) shows that  the conditional probability to exit on $t$, given an entrance on $s$, is $w_{st}=n^0_{st}/n^0_{s\bullet}=I(s<t)\, d_t/D_s$.

*** ici la (les) figure de l'exemple  ``starting from Maladière, Riant-Cour, Dapples"... ? ***

*** ici l'équivalence avec l'approche chaîne de Markov de Guillaume ? ***

>>>>

Let  $i=1,\ldots, l$ enumerate the bus stops in increasing order,  i.e. $F(i)=i+1$. The initial prior is simply $g_{st}=\frac{I(s<t)}{(l-1)(l-2)}$ (where $I(.)$ denotes the 0/1  indicator function), capturing solely the unidirectional nature of trips. The margins of the empirical distribution $f_{st}$, as well as the total flow, are here known : 
\begin{displaymath}
\alpha_s=\frac{a_s}{a_\bullet}\qquad\qquad\qquad \beta_t=\frac{b_t}{b_\bullet}\qquad\qquad\qquad n_{\bullet\bullet}=a_{\bullet}=b_\bullet
\end{displaymath}
and (\ref{Sol}) yields, after reparameterization, maximum entropy flows of the form
\begin{equation}
\label{nosignle}
n_{st}=I(s<t)\, c_s\, d_t \qquad\qquad\mbox{where}\quad \sum_{s}c_s\, \sum_{t>s}d_t=n_{\bullet\bullet}
\end{equation}
where (setting $D_s:=\sum_{t>s}d_t$ and $C_t:=\sum_{s<t}c_s$) the constraints (\ref{Sol_LagMult}) equivalently express as 
\begin{equation}
\label{dis embarking constraints}
c_s=\frac{a_s}{\sum_{t>s}d_t}=\frac{a_s}{D_s}
\qquad\qquad\qquad
d_t=\frac{b_t}{\sum_{s<t}c_s}=\frac{b_t}{C_t}
\end{equation}
to be solved by iterative fitting.

Interestingly enough, the form (\ref{nosignle}) for the flows is reminiscent of the {\em gravity flows} of quantitative Geography \cite{wilson1967statistical}  \cite{erlander1990gravity} \cite{bavaud2002quasi}, where 
$c_s$ is the {\em push factor}, $d_t$ is the {\em pull factor}, and $I(s<t)$ the {\em distance deterrence function}. Yet, instead of  
being symmetric in $s,t$ and decreasing with the distance $|s-t|$, the distance deterrence function is here asymmetric due to the line orientation, but otherwise constant.

This constancy entails the following Markovian behaviour for flows: let $m_{st}$ be the number of travelers embarking at stop $s$ and still inside the bus at stop $t>s$, and let $\rho_{st}$ the probability that travelers embarking at $s$ will disembark at $t$. By (\ref{nosignle}), 
\begin{displaymath}
m_{st}=\sum_{u\ge t}n_{su}=c_s \sum_{u\ge t} I(s<u)\, d_u =c_s (d_t+D_t) 
\end{displaymath}
The empirical estimate of $\rho_{st}$ is given by the proportion, among the travelers embarking at $s$ and 
 still present at $t>s$,  of travelers disembarking at $t$, that is 
\begin{displaymath}
\rho_{st}=\frac{n_{st}}{m_{st}}=\frac{c_s\,  d_t}{c_s\,  (d_t+D_t)}=\frac{d_t}{d_t+D_t}=\frac{b_t}{C_t\, (d_t+D_t)}\le 1
\end{displaymath}
which depends on $t$ only: it  finally  turns out that the disembarkment probability $\rho_t=\frac{d_t}{d_t+D_t}$ at $t$ is {\em independent} of the embarkment stop $s$. Said otherwise, a traveler embarking at any stop $s$ (and thus necessarily in the bus at $F(s)=s+1$) experiences the {\em same disembarkment probability} at each  stop $t$.

This Markov property, enjoyed by maximum-entropic flows, contrasts other possible solutions, such as  the ``first in, first out" (FIFO) flows (homogenizing the 
traveled distances among users) or the ``last in, first out" (LIFO) flows (tending to generate maximally  contrasted traveled distances).

@@

***  @G, ici davantage de développements sur l'approche Markovienne, ainsi que le joli schéma ?  ***

@@

*** @G, @R, ici la (les) figure de l'exemple  ``starting from Maladière, Riant-Cour, Dapples"... ? ***
